---
title: "**MATH 3190 Homework 8**"
subtitle: |
  Melanie Jensen\
  Due April 2, 2026\
  Focus: Notes 9\
  Total Points: 90 + 10 (Self-Critique Comment and Uploading to GitHub) = 100
format: 
  html:
    embed-resources: true # Keep this so the html file is easily shareable.
urlcolor: blue
execute: # Will set these as the default for all code chunks
  eval: true
  echo: true
  cache: true
editor_options: 
  chunk_output_type: console
---

Note that your homework should be completed in Quarto. You can just add your answers and code to this document in the appropriate part. The file should be rendered as an html document or pdf document (html is quicker to render and you will need an installation of LaTeX to render as a pdf). You will "turn in" this homework by uploading the files, the .qmd and output file, to your GitHub `Math_3190_Assignment` repository in the `Homework/Homework_8` directory. Make sure all supporting files (if there are any) are in this directory as well. You should use `git` and `Unix` commands in a terminal for this (i.e. don't manually upload files to GitHub). In addition, you will submit just your html or pdf file on Canvas to unlock the solutions.

Don't forget to also put a [self-critique comment]{.underline} on Canvas indicating what you did correctly and incorrectly and include any questions you may have about the assignment after submitting the assignment and thoroughly looking through the solutions.



# Problem 1 (10 points)
Three airlines serve a small town in Ohio. Airline A has 52% of all scheduled flights, airline B has 35% and airline C has the remaining 13%. Their on-time rates are 85%, 67%, and 41%, respectively. A flight just left on-time. What is the probability that it was a flight of airline A?

```{r}
#| label: p1

# P(A | on-time) = ( P(on-time | A) * P(A) ) / P(on-time)

print((0.85 * 0.52) / (0.85 * 0.52 + 0.67 * 0.35 + 0.41 * 0.13))

```

# Problem 2 (15 points)
Suppose we have a data set with each observation $x_i$ independent and identically exponentially distributed for $i=1,2,\dots,n$. That is, $x_i\sim \text{Exp}(\lambda)$ where $\lambda$ is the rate parameter. We would like to find a posterior (or at least a function proportional to it) for $\lambda$.

### Part a (6 points)
Write down the likelihood function (or a function proportional to it) in this situation. We would call this $p(x|\lambda)$.

$$p(x | \lambda) = \prod_{i=1}^{n} \lambda e^{-\lambda x_i} = \lambda^n \prod_{i=1}^{n} e^{-\lambda x_i} = \lambda^n e^{-\lambda \sum_{i=1}^n x_i}$$

### Part b (6 points)
Now let $\lambda$ have a normal prior with mean 0.1 and variance 1: $\lambda\sim N(1/10,\ 1)$. Use this and the likelihood from part (a) to write down a function that is proportional to the posterior of $\lambda$ given ${x}$. We call this $p(\lambda|{x})$.

$$p(\lambda \mid x) \propto \lambda^n \exp\!\left\{-\lambda \sum_{i=1}^n x_i - \frac{(\lambda - 0.1)^2}{2}\right\}$$

### Part c (3 points)
Which would be more appropriate here to obtain samples of $\lambda$, the Gibbs or Metropolis algorithm? Explain why. You may want to look on page 171 of Notes 9 in the conjugate prior table.

It would be more appropriate to use the Metropolis algorithm here since the posterior of $\lambda$ does not have a known distribution that we can easily sample from using the Gibbs algorithm. Additionally, as we know the function proportional to the posterior, we can use the Metropolis algorithm to sample from it much more easily. 

# Problem 3 (30 points)
Suppose we have the vector
`x = c(1.83, 1.72, 2.13, 2.49, 0.90, 2.01, 1.51, 3.12, 1.29, 1.54, 2.94, 3.02, 0.93, 2.78)` that we believe comes from a gamma distribution with shape of 10 and some rate $\beta$:
$x_i\sim\text{Gam}(10,\ \beta)$. We will use sampling to obtain some information about $\beta$. Let’s put a gamma prior on $\beta$ with a shape of $\alpha_0$ and a rate of 1: $\beta\sim\text{Gam}(\alpha_0,\ 1)$.

### Part a (5 points)
Use the fact that this is a conjugate prior to write down what kind of distribution the posterior of $\beta$, which is $p(\beta|{x})$, is. 

$x_i \sim \text{Gam}(10, \beta)$

$$f(x_i \mid \beta) = \frac{\beta^{10}}{\Gamma(10)} x_i^{9} e^{-\beta x_i}$$

Likelihood:

$$p({x} \mid \beta) = \prod_{i=1}^n \frac{\beta^{10}}{\Gamma(10)} x_i^{9} e^{-\beta x_i} = \frac{\beta^{10n}}{\Gamma(10)^n} \left(\prod_{i=1}^n x_i\right)^9 e^{-\beta \left(\sum_{i=1}^n x_i\right)}$$

$$p({x} \mid \beta) \propto \beta^{10n} e^{-\beta \left(\sum_{i=1}^n x_i\right)}$$

Prior:

$$p(\beta) = \frac{1^{\alpha_0}}{\Gamma(\alpha_0)} \beta^{\alpha_0 - 1} e^{-\beta}$$

$$p(\beta) \propto \beta^{\alpha_0 - 1} e^{-\beta}$$

Posterior = Likelihood * Prior:

$$p(\beta \mid {x}) \propto p({x} \mid \beta) \cdot p(\beta)$$

$$p(\beta \mid {x}) \propto \left[\beta^{10n} e^{-\beta \left(\sum_{i=1}^n x_i\right)}\right] \cdot \left[\beta^{\alpha_0 - 1} e^{-\beta}\right]$$

$$p(\beta \mid {x}) \propto \beta^{10n + \alpha_0 - 1} \cdot e^{-\beta\left(\left(\sum_{i=1}^n x_i\right) + 1\right)}$$

$$p(\beta \mid {x}) = \text{Gam}\!\left(10n + \alpha_0,\ \left(\sum_{i=1}^n x_i\right) + 1\right)$$

### Part b (5 points)
Let $\alpha_0=1$. In an **R** code chunk, sample 10,000 $\beta$ values from the distribution you wrote down in part (a) using the `rgamma()` function and report the 95% credible interval for $\beta$ using the 2.5th and 97.5th percentiles. 

```{r}
#| label: p3b

x = c(1.83, 1.72, 2.13, 2.49, 0.90, 2.01, 1.51, 3.12, 1.29, 1.54, 2.94, 3.02, 0.93, 2.78)
alpha_0 <- 1

set.seed(2026)
sample_betas <- rgamma(10000, shape = 10 * length(x) + alpha_0, rate = sum(x) + 1)

quantile(sample_betas, probs = c(0.025, 0.975))

```

### Part c (3 points)
Repeat part (b) with $\alpha_0 = 10$.

```{r}
#| label: p3c

x = c(1.83, 1.72, 2.13, 2.49, 0.90, 2.01, 1.51, 3.12, 1.29, 1.54, 2.94, 3.02, 0.93, 2.78)
alpha_0 <- 10

set.seed(2026)
sample_betas <- rgamma(10000, shape = 10 * length(x) + alpha_0, rate = sum(x) + 1)

quantile(sample_betas, probs = c(0.025, 0.975))

```

### Part d (3 points)
Repeat part (b) with $\alpha_0 = 100$.

```{r}
#| label: p3d

x = c(1.83, 1.72, 2.13, 2.49, 0.90, 2.01, 1.51, 3.12, 1.29, 1.54, 2.94, 3.02, 0.93, 2.78)
alpha_0 <- 100

set.seed(2026)
sample_betas <- rgamma(10000, shape = 10 * length(x) + alpha_0, rate = sum(x) + 1)

quantile(sample_betas, probs = c(0.025, 0.975))

```

### Part e (4 points)
Comment on the effect the value of $\alpha_0$ has on the credible intervals for $\beta$ as revealed in parts (b), (c), and (d).

```{r}
#| label: p3e

print(5.644936 - 4.057098)
print(5.977442 - 4.340462)
print(9.271471 - 7.204932)

```

As alpha_0 increases, the intervals not only shift to the right, but they are also getting wider. This makes sense because as alpha_0 increases, the shape parameter and the variance of the prior distribution for beta are increasing, which puts more weight on larger values of betas while also increasing the variance of the posterior distribution for beta. This results in wider posterior probability intervals for beta as we say in the random sample outputs above. 

### Part f (6 points)
Now suppose we have twice as much data (given in the **R** code chunk below). Repeat parts (b), (c), and (d) using this `x` vector given below instead of the other one and report the three 95% credible intervals. Note, this new vector x will change the shape and rate parameters used in the `rgamma()` functions.
```{r}
#| label: new_x
x <- c(1.83, 1.72, 2.13, 2.49, 0.90, 2.01, 1.51, 3.12, 1.29, 1.54,
       2.94, 3.02, 0.93, 2.78, 2.76, 1.70, 1.42, 2.16, 1.07, 2.21,
       2.38, 2.27, 1.72, 1.44, 1.54, 1.72, 1.87, 1.39)

alpha_0 <- 1

set.seed(2026)
sample_betas <- rgamma(10000, shape = 10 * length(x) + alpha_0, rate = sum(x) + 1)

quantile(sample_betas, probs = c(0.025, 0.975))

alpha_0 <- 10

set.seed(2026)
sample_betas <- rgamma(10000, shape = 10 * length(x) + alpha_0, rate = sum(x) + 1)

quantile(sample_betas, probs = c(0.025, 0.975))

alpha_0 <- 100

set.seed(2026)
sample_betas <- rgamma(10000, shape = 10 * length(x) + alpha_0, rate = sum(x) + 1)

quantile(sample_betas, probs = c(0.025, 0.975))

```

### Part g (4 points)
In this problem, the true $\beta$ value is 5. Write a sentence or two about the effect adding more data has to these credible intervals by comparing the intervals from parts (b)-(d) to the intervals from part (f).

```{r}
#| label: p3g

print(5.729326 - 4.538663)
print(5.902571 - 4.693303)
print(7.630896 - 6.246754)

```

We can see for all of the intervals, adding more data narrowed the posterior probability intervals. While adding more data also reduced how much the intervals shift to the right, we can see that when alpha_0 is increased up to 100, both intervals still fail to capture the true beta of 5, with the interval found in part f (more data) being closer to capturing the true beta than the interval found in part d. 

# Problem 4 (35 points)
Let’s apply the Bayesian framework to a regression problem. In the `math3190package` library is the `treeseeds` dataset (it is also on GitHub) that contains information about species of tree, the count of seeds it produces, and the average weight of those seeds in mg.

### Part a (3 points)
Using the `treeseeds` data, create new variables defined by taking the log of the counts and weights. Fit an OLS regression model using logged weight to predict logged count.

```{r}
#| label: treeseeds
library(math3190package) |> suppressPackageStartupMessages()
tree <- treeseeds |>
  mutate(lcount = log(count),
         lweight = log(weight))

ols_model <- lm(lcount ~ lweight, data = tree)
summary(ols_model)

```

### Part b (5 points)
Please read this carefully and try to understand what is being done here. There is just one question at the end of this part for you. We will walk through the mathematics and code of obtaining the posterior and samples together here since this problem will focus on looking at diagnostics, but I want you to see the **Metropolis algorithm** in action! Make sure to read this in the html or pdf file where all equations are rendered nicely. 

Assuming the true errors are normal with mean 0 and variance $\sigma^2$, $\epsilon_i\sim N(0,\ \sigma^2)$, it can be shown that each $y_i$ has the distribution
$$
p(y_i|x_i,\beta_0,\beta_1\sigma^2) = \frac{1}{\sqrt{2\pi\sigma^2}}\exp\left(
-\frac{1}{2\sigma^2}(y_i-\beta_0-\beta_1x_i)^2\right)
$$

So, we can write the likelihood is 
$$
p(y_i|x_i,\beta_0,\beta_1\sigma^2)\propto\exp\left(
-\frac{1}{2\sigma^2}(y_i-\beta_0-\beta_1x_i)^2\right)
$$
where $y_i$ is the log(count) for observation $i$ and $x_i$ is the log(weight) for observation $i$. Note that here we think of $\boldsymbol{y}$ as being random and ${x}$ as being fixed. We could, in theory, think of the vector ${x}$ as also being random and put a prior on it. But we won’t do that here.

Now, let’s just put uniform priors on $\beta_0$ and $\beta_1$ so the priors are proportional to 1. Also, let’s assume $\sigma^2=1$. This seems reasonable since $s_e^2$, the MSE, is 0.877. Of course, we could put a prior on $\sigma^2$ as well and sample it too, but we will focus on only sampling $\beta_0$ and $\beta_1$.

Now, with those uniform priors, and plugging in 1 for $\sigma^2$, we have that the joint posterior of $\beta_0$ and $\beta_1$ is:
$$
p(\beta_0,\beta_1|{x},\boldsymbol{y}) \propto\exp\left(
-\frac{1}{2}\sum_{i=1}^n(y_i-\beta_0-\beta_1x_i)^2\right)=f(\beta_0,\beta_1|{x},\boldsymbol{y}).
$$

Then, we can take the log to get 
$$
\ln(f(\beta_0,\beta_1|{x},\boldsymbol{y}))=-\frac{1}{2}\sum_{i=1}^n(y_i-\beta_0-\beta_1x_i)^2.
$$

Our goal now is to obtain samples of $\beta_0$ and $\beta_1$. Let’s use the Metropolis algorithm to do this. Using the log of the function proportional to the joint posterior of $\beta_0$ and $\beta_1$, $\ln\Big(f(\beta_0,\beta_1|{x},\boldsymbol{y})\Big)$, we'll write a Metropolis algorithm in **R**. For $\beta_0$, we will use a normal proposal distribution centered at the previous value, $\beta_0^{(i)}$, with a standard deviation of 0.8 and for $\beta_1$, and we will use a normal proposal distribution centered at the previous value, $\beta_1^{(i)}$, with a standard deviation of 0.1. The starting values don’t matter too much, but we'll use $\beta_0^{(0)}=10$ and $\beta_1^{(0)}=-0.5$.

This code will generate 10,000 samples and plot the chains for $\beta_0$ and $\beta_1$. In addition, the 95% credible interval for $\beta_0$ and $\beta_1$ are given.

```{r}
#| label: mcmc
#| fig-height: 6
#| fig-width: 8
library(patchwork)
set.seed(2026)
nsamps <- 10000
beta0 <- rep(0, nsamps)
beta1 <- rep(0, nsamps)
beta0[1] <- 10
beta1[1] <- -0.5
for (i in 1:(nsamps - 1)) {
  beta0_star <- rnorm(1, beta0[i], 0.8) # Proposed new value for intercept
  beta1_star <- rnorm(1, beta1[i], 0.1) # Proposed new value for slope
  # y is tree$lcount and x is tree$lweight
  lf1 <- -1 / 2 * sum((tree$lcount - beta0_star - beta1_star * tree$lweight)^2)
  lf2 <- -1 / 2 * sum((tree$lcount - beta0[i] - beta1[i] * tree$lweight)^2)
  if (log(runif(1)) < lf1 - lf2) {
    # Accept the proposed values as new sample values if the above 
    #   condition is satisfied
    beta0[i + 1] <- beta0_star
    beta1[i + 1] <- beta1_star
  } else {
    # Otherwise set the new sample values to be equal to the previous ones
    beta0[i + 1] <- beta0[i]
    beta1[i + 1] <- beta1[i]
  }
}

mcmc_df <- data.frame(index = 1:nsamps, beta0 = beta0, beta1 = beta1)
mcmc_plot1 <- 
  ggplot(mcmc_df, aes(x = index, y = beta0)) +
  geom_line() +
  theme_test() +
  scale_x_continuous(breaks = seq(0, 10000, by = 2000)) +
  scale_y_continuous(breaks = seq(8, 11, by = 0.5)) + 
  labs(x = "Index", y = expression(beta[0]))

mcmc_plot2 <- 
  ggplot(mcmc_df, aes(x = index, y = beta1)) +
  geom_line() +
  theme_test() +
  scale_x_continuous(breaks = seq(0, 10000, by = 2000)) +
  scale_y_continuous(breaks = seq(-0.75, -0.25, by = 0.1)) + 
  labs(x = "Index", y = expression(beta[1]))

mcmc_plot1 / mcmc_plot2

quantile(beta0, c(.025, .975))
quantile(beta1, c(.025, .975))
```

The plots of the chains are given in the output above. 

The 95% credible interval for $\beta_0$ is $(8.53, 10.21)$.

The 95% credible interval for $\beta_1$ is $(-0.661, -0.368)$.

Your intervals may differ a bit due to the random sampling.

Okay, here is the question for you! Based on the plots of the chains above, does it look like the Metropolis sampling worked fairly well? Explain. 

It looks like the Metropolis sampling worked fairly well as the sampling chains for both $\beta_0$ and $\beta_1$ do not appear to be too slow or get stuck in place for too long, which indicates that the proposed distributions for both $\beta_0$ and $\beta_1$ fit the target distribution well. 

### Part c (3 points)
Interpret both of the credible intervals from part (b). 

There is a 95% posterior probability that the true value of $\beta_0$ is between 8.5264 and 10.2106. 

There is a 95% posterior probability that the true value of $\beta_1$ is between -0.6607 and -0.3681.

### Part d (5 points)
Find and report the integrated autocorrelation time for the $\beta_0$ and $\beta_1$ chains. Each chain will have their own $\hat{\tau}_{int}$ value, so you should report two (although they will be similar). 

```{r}
#| label: p4d

tau_beta0 <- 1 + 2 * sum(abs(acf(beta0, lag.max = 100, plot = FALSE)$acf))
tau_beta1 <- 1 + 2 * sum(abs(acf(beta1, lag.max = 100, plot = FALSE)$acf))

tau_beta0
tau_beta1

```

### Part e (3 points)
Based on the integrated autocorrelation time for the $\beta_0$ and $\beta_1$ chains, how many MCMC samples would you need to generate to get the equivalent of 10,000 independent samples?

The integrated autocorrelation of $\beta_0$ is $\hat{\tau}_{int} = 25.73$ 
and for $\beta_1$ is $\hat{\tau}_{int} = 26.62$. This means that approximately every 27 samples, we obtain one effectively independent sample. To get the equivalent of 10,000 independent samples, we would need to generate approximately 266,617 samples.

### Part f (3 points)
Let’s compare these credible intervals to some other intervals. First, obtain the 95%  $t$ confidence intervals for $\beta_0$ and $\beta_1$ just using the `confint()` function on the model you fit in part (a) and report them here.

```{r}
#| label: p4f

confint(ols_model, level = 0.95)

```

### Part g (8 points)
Now let's obtain confidence intervals using bootstrapping the cases like we did with in Notes 7 and HW 5. You can use the `bootstrapping()` function in the `math3190package` for this or do it yourself with a loop. Set a seed and then using at least 10,000 bootstrap samples, report the 95% percentile confidence intervals for $\beta_0$ and $\beta_1$.

```{r}
#| label: p4g

coef_func <- function(data, i) {
  fit <- lm(lcount ~ lweight, data = data[i, ])
  coef(fit)
}

set.seed(2026)
boot_out <- bootstrapping(tree, coef_func, 10000)

# Beta 0
quantile(boot_out$t[, 1], c(0.025, 0.975))

# Beta 1
quantile(boot_out$t[, 2], c(0.025, 0.975))

```

### Part i (5 points)
Write a couple sentences comparing all of the intervals in parts (b), (f), and (g).


```{r}
#| label: p4i

library(knitr)

interval_table <- data.frame(
  Method = c("OLS confint()", "Metropolis", "Bootstrap"),
  Beta0_2.5 = c(8.53, 8.53, 8.396988),
  Beta0_97.5 = c(10.21, 10.21, 10.289586),
  Beta1_2.5 = c(-0.661, -0.661, -0.6671630),
  Beta1_97.5 = c(-0.368, -0.368, -0.3496652)
)

kable(interval_table, 
      col.names = c("Method", "$\\beta_0$ 2.5%", "$\\beta_0$ 97.5%", 
                    "$\\beta_1$ 2.5%", "$\\beta_1$ 97.5%"),
      digits = 4,
      caption = "95% Confidence/Credible Intervals for $\\beta_0$ and $\\beta_1$")
```

We can see that across the board all of the intervals are very similar. This shows that all of the methods are identifying the same information about the parameters. This indicates that the prior on the Metropolis sampling did not have a large influence on the posterior, which makes the likelihood the main driver. This is why the Metropolis and the OLS intervals are basically the same. The bootstrap intervals, while still being very similar, are a bit wider than the other two methods, which is likely due to the fact that there are not as many assumptions being made with bootstrapping as there are with the other two methods, which leads to more uncertainty in the estimates and thus wider intervals.