---
title: "**MATH 3190 Homework 4**"
subtitle: |
  Due February 19, 2026\
  Focus: Notes 6\
  Name: Melanie Jensen\
  Total Points: 90 + 10 (Self-Critique Comment and Uploading to GitHub) = 100
format: 
  html:
    embed-resources: true # Keep this so the html file is easily shareable.
urlcolor: blue
engine: r
execute: # Will set these as the default for all code chunks
  message: false
  warning: false
  eval: true
  echo: true
  cache: true
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
---

Note that your homework should be completed in Quarto. You can just add
your answers and code to this document in the appropriate part. The file
should be rendered as an html document or pdf document (html is quicker
to render and you will need an installation of LaTeX to render as a
pdf). You will "turn in" this homework by uploading the files, the .qmd
and output file, to your GitHub `Math_3190_Assignment` repository in the
`Homework/Homework_4` directory. Make sure all supporting files (if
there are any) are in this directory as well. You should use `git` and
`Unix` commands in a terminal for this (i.e. don't manually upload files
to GitHub). In addition, you will submit just your html or pdf file on
Canvas to unlock the solutions.

Don't forget to also put a [self-critique comment]{.underline} on Canvas
indicating what you did correctly and incorrectly and include any
questions you may have about the assignment after submitting the
assignment and thoroughly looking through the solutions.

Some of the parts in this homework, like 1a, 2b, 2c, 3a, and 3b require
writing down some math-heavy expressions. You may either type it up
using LaTeX-style formatting in Quarto (preferred), or you can write it
by hand (neatly) and include pictures or scans of your work in your
Quarto document.

# Problem 1 (33 points)

Suppose we want to find the value of $x$ and the objection function
value for finding the global maximum and minimum of the function
$f(x)=\ln(1+x^2)(1+x)e^{-x^2}$. We will implement gradient descent and
Newton's method for finding these extrema.

### Part a (5 points)

Find $f'(x)$ for this function. Do this by hand as a nice derivative
refresher. You don't need to include all of your work, though. Feel free
to check your answer using something like Wolfram\|Alpha. Write down the
derivative here.

$$
f(x) = e^{-x^2}ln(1+x^2) + xe^{-x^2}ln(1+x^2)
$$

a.  $e^{-x^2}ln(1+x^2)$\
    $\begin{alignedat}{2}
     u &= e^{-x^2} & \qquad du &= -2xe^{-x^2} \\
     v &= ln(1+x^2) & \qquad dv &= \frac{2x}{1+x^2} \\
    \end{alignedat}$

b.  $xe^{-x^2}ln(1+x^2)$\
    $\begin{alignedat}{2}
     u &= xe^{-x^2} & \qquad du &= (1-2x^2)e^{-x^2} \\
     v &= ln(1+x^2) & \qquad dv &= \frac{2x}{x^2+1} \\
    \end{alignedat}$

$$
f'(x) = (e^{-x^2})(\frac{2x}{1+x^2}) + (-2xe^{-x^2})(ln(1+x^2)) + (xe^{-x^2})(\frac{2x}{x^2+1}) + ((1-2x^2)e^{-x^2})(ln(1+x^2))
$$

$$
f'(x) = (-2x^2 - 2x + 1)e^{-x^2}ln(1+x^2) + \frac{2x(1+x)e^{-x^2}}{1+x^2}
$$

### Part b (4 points)

Below is the code of a function that implements gradient descent using a
backtracking line search each iteration to find $\gamma_k$: the step
size. This function takes eight inputs:

-   The starting value (or vector)
-   The function to optimize
-   The function's derivative (or gradient)
-   A logical indicating if we want to find a max with a default of
    `FALSE`.
-   The maximum number of iterations with a default of 200
-   The initial step size with a default of 1
-   The line search beta parameter with a default of 0.5.
-   The stopping tolerance with a default of `1e-10`.

The output of this function is a list with the $x$ value (or vector)
that optimizes the function, the function value at that point, and the
number of iterations it took to converge. This function works in the
case of one dimension or multiple dimensions since you'll use it again
in problem 2.

Take a few minutes to study this function to make sure you understand
how it works and what it is doing. Write a few sentences explaining how
this function works.

```{r}
#| label: backtracking_function
#| eval: true
#| echo: true

gradient_desc <- function(starting, f, gradient, max = F, maxit = 200, step_size = 1, line_parameter = 0.5, tol = 1e-10) {

  if(max == T) {
    obj_func <- function(x) {
      -f(x)
    }
    func_deriv <- function(x) {
      -gradient(x)
    }
  } else {
    obj_func <- f
    func_deriv <- gradient
  }
  xk <- starting
  for(k in 1:maxit) {
    gamma <- step_size
    line_beta <- line_parameter
    # Below is the backtracking line search
    while(obj_func(xk - gamma * func_deriv(xk)) > 
          obj_func(xk) - gamma/4 * sum(func_deriv(xk)^2)) {
      gamma <- line_beta * gamma
    }
    # Once we have a good step size, continue
    xnew <- xk - gamma * func_deriv(xk)
    if(norm(xk - xnew, type = "2") < tol) {
      break
    }
    xk <- xnew
    }
  return(list(x = xk, function_val = f(xk), iters = k))
}
```

First, the function determines whether the user wants to minimize or
maximize the objective function. If the user wants to maximize the
objective function, the function will minimize the negative objective
function. Then, the function performs backtracking line search. This
algorithm starts with an initial step size and then checks if the
function value at the new point (after taking a step in the direction of
the negative gradient) is sufficiently smaller than the function value
at the current point. If it is not, the step size is reduced by
multiplying it by a line parameter (which is between 0 and 1). This
process continues until a suitable step size is found. After finding a
good step size, the function updates the current point and checks for
convergence by comparing the distance between the new point and the
current point to a specified tolerance. If convergence is achieved, the
loop breaks, and the function returns the optimal point, the function
value at that point, and the number of iterations taken to converge.

### Part c (5 points)

Use the gradient descent function to find the global **min** of
$f(x)=\ln(1+x^2)(1+x)e^{-x^2}$. Try using several starting points. Keep
track of and report the number of iterations needed to converge at the
different starting points.

```{r}
#| label: gradient descent min example

library(Deriv)
library(ggplot2)

f <- function(x) {
  log(1+x^2)*(1+x)*exp(-x^2)
}
f_prime <- function(x) {
  (-2*x^2-2*x+1)*exp(-x^2)*log(1+x^2)+(2*x*(1+x)*exp(-x^2))/(1+x^2)
}

ggplot(data.frame(x = c(-5, 5)), aes(x = x)) +
  stat_function(fun = f, color = "royalblue", size = 1) +
  
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5) +

  labs(title = expression(paste(f(x) == ln(1+x^2)(1+x)*e^{-x^2})),
       x = "x", y = "f(x)") +
  theme_minimal()

gradient_desc(starting = -3, f = f, gradient = f_prime)
gradient_desc(starting = -2, f = f, gradient = f_prime)
gradient_desc(starting = -1, f = f, gradient = f_prime)
gradient_desc(starting = 0, f = f, gradient = f_prime)
gradient_desc(starting = 1, f = f, gradient = f_prime)
gradient_desc(starting = 2, f = f, gradient = f_prime)
gradient_desc(starting = 3, f = f, gradient = f_prime)

```

The approximate minimum occurs at $x = -1.469439$ with a function value
of $-0.06232362$. Even with a starting point of $x = -1$ it still took
$31$ iterations to converge.

### Part d (4 points)

Use the gradient descent function to find the global **max** of
$f(x)=\ln(1+x^2)(1+x)e^{-x^2}$. Again, try several starting points. Keep
track of and report the number of iterations needed to converge at the
different starting points.

```{r}
#| label: gradient descent max example

f <- function(x) {
  log(1+x^2)*(1+x)*exp(-x^2)
}
f_prime <- function(x) {
  (-2*x^2-2*x+1)*exp(-x^2)*log(1+x^2)+(2*x*(1+x)*exp(-x^2))/(1+x^2)
}

ggplot(data.frame(x = c(-5, 5)), aes(x = x)) +
  stat_function(fun = f, color = "royalblue", size = 1) +
  
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5) +

  labs(title = expression(paste(f(x) == ln(1+x^2)(1+x)*e^{-x^2})),
       x = "x", y = "f(x)") +
  theme_minimal()

gradient_desc(starting = -3, f = f, gradient = f_prime, max = T)
gradient_desc(starting = -2, f = f, gradient = f_prime, max = T)
gradient_desc(starting = -1, f = f, gradient = f_prime, max = T)
gradient_desc(starting = 0, f = f, gradient = f_prime, max = T)
gradient_desc(starting = 1, f = f, gradient = f_prime, max = T)
gradient_desc(starting = 2, f = f, gradient = f_prime, max = T)
gradient_desc(starting = 3, f = f, gradient = f_prime, max = T)

```

The approximate maximum occurs at $x = 0.9868668$ with a function value
of $0.5101816$. Even with a starting value of 1, it still took $22$
iterations to converge.

### Part e (8 points)

Use the fact that $$
f''(x)=\dfrac{2e^{-x^2}}{(1+x^2)^2}\bigg((2x^3+2x^2-3x-1)(1+x^2)^2\ln(1+x^2)-4x^5-4x^4-3x^3-5x^2+3x+1\bigg)
$$

to implement Newton's method to find the global max **and** the global
min. Use the same starting points you did in parts c and d. Keep track
of and report the number of iterations needed to converge at the
different starting points. And yes, that second derivative is very
messy! This is one reason why Newton's method is less popular. You don't
have to write a function for this part to implement Newton's method, but
you can if you'd like.

```{r}
#| label: newton's method min and max

library(tidyverse)

# Max

f <- function(x) {
  log(1+x^2)*(1+x)*exp(-x^2)
}
f_prime <- function(x) {
  (-2*x^2-2*x+1)*exp(-x^2)*log(1+x^2)+(2*x*(1+x)*exp(-x^2))/(1+x^2)
}
f_double_prime <- function(x) {
  ((2*exp(-x^2))/((1+x^2)^2))*(((2*x^3+x*x^2-3*x-1)*(1+x^2)^2)*log(1+x^2)-
                                 4*x^5-4*x^4-3*x^3-5*x^2+3*x+1)
}

maxit <- 200
betak <- c(-3, -2, -1, 0, 1, 2, 3)
found_roots <- c()
iters <- c()

for (i in betak) {
  x <- i
  for (k in 1:maxit) {
    x_next <- x - f_prime(x) / f_double_prime(x)
    if (abs(x_next - x) < 1e-10) {
      break}
    x <- x_next}
  found_roots <- c(found_roots, x)
  iters <- c(iters, k)}
n_max_results <- tibble(
  start_val = betak,
  x = found_roots,
  function_val = round(f(found_roots),10),
  iterations = iters)

# Min

found_roots <- c()
iters <- c()

for (i in betak) {
  x <- i
  for (k in 1:maxit) {
    x_next <- x - f_prime(x) / abs(f_double_prime(x))
    if (abs(x_next - x) < 1e-10) {
      break}
    x <- x_next}
  found_roots <- c(found_roots, x)
  iters <- c(iters, k)}
n_min_results <- tibble(
  start_val = betak,
  x = found_roots,
  function_val = round(f(found_roots),10),
  iterations = iters)

```

### Part f (3 points)

Compare the number of iterations needed for convergence for gradient
descent and for Newton's method.

```{r}
#| label: comparison

g_min_results <- tibble(
  start_val = betak,
  x = c(gradient_desc(starting = -3, f = f, gradient = f_prime)$x, 
        gradient_desc(starting = -2, f = f, gradient = f_prime)$x, 
        gradient_desc(starting = -1, f = f, gradient = f_prime)$x, 
        gradient_desc(starting = 0, f = f, gradient = f_prime)$x, 
        gradient_desc(starting = 1, f = f, gradient = f_prime)$x, 
        gradient_desc(starting = 2, f = f, gradient = f_prime)$x, 
        gradient_desc(starting = 3, f = f, gradient = f_prime)$x),
  function_val = c(gradient_desc(starting = -3, f = f, 
                                 gradient = f_prime)$function_val, 
                   gradient_desc(starting = -2, f = f, 
                                 gradient = f_prime)$function_val, 
                   gradient_desc(starting = -1, f = f, 
                                 gradient = f_prime)$function_val, 
                   gradient_desc(starting = 0, f = f, gradient = f_prime)$function_val, 
                   gradient_desc(starting = 1, f = f, gradient = f_prime)$function_val, 
                   gradient_desc(starting = 2, f = f, gradient = f_prime)$function_val, 
                   gradient_desc(starting = 3, f = f, gradient = f_prime)$function_val), 
  iterations = c(gradient_desc(starting = -3, f = f, gradient = f_prime)$iters, 
                 gradient_desc(starting = -2, f = f, gradient = f_prime)$iters, 
                 gradient_desc(starting = -1, f = f, gradient = f_prime)$iters, 
                 gradient_desc(starting = 0, f = f, gradient = f_prime)$iters, 
                 gradient_desc(starting = 1, f = f, gradient = f_prime)$iters, 
                 gradient_desc(starting = 2, f = f, gradient = f_prime)$iters, 
                 gradient_desc(starting = 3, f = f, gradient = f_prime)$iters)
)

g_max_results <- tibble(
  start_val = betak,
  x = c(gradient_desc(starting = -3, f = f, gradient = f_prime, max = T)$x, 
        gradient_desc(starting = -2, f = f, gradient = f_prime, max = T)$x, 
        gradient_desc(starting = -1, f = f, gradient = f_prime, max = T)$x, 
        gradient_desc(starting = 0, f = f, gradient = f_prime, max = T)$x, 
        gradient_desc(starting = 1, f = f, gradient = f_prime, max = T)$x, 
        gradient_desc(starting = 2, f = f, gradient = f_prime, max = T)$x, 
        gradient_desc(starting = 3, f = f, gradient = f_prime, max = T)$x),
  function_val = c(gradient_desc(starting = -3, f = f, gradient = f_prime, max = T)$function_val, 
                   gradient_desc(starting = -2, f = f, gradient = f_prime, max = T)$function_val, 
                   gradient_desc(starting = -1, f = f, gradient = f_prime, max = T)$function_val, 
                   gradient_desc(starting = 0, f = f, gradient = f_prime, max = T)$function_val, 
                   gradient_desc(starting = 1, f = f, gradient = f_prime, max = T)$function_val, 
                   gradient_desc(starting = 2, f = f, gradient = f_prime, max = T)$function_val, 
                   gradient_desc(starting = 3, f = f, gradient = f_prime, max = T)$function_val),
  iterations = c(gradient_desc(starting = -3, f = f, gradient = f_prime, max = T)$iters, 
                 gradient_desc(starting = -2, f = f, gradient = f_prime, max = T)$iters, 
                 gradient_desc(starting = -1, f = f, gradient = f_prime, max = T)$iters, 
                 gradient_desc(starting = 0, f = f, gradient = f_prime, max = T)$iters, 
                 gradient_desc(starting = 1, f = f, gradient = f_prime, max = T)$iters, 
                 gradient_desc(starting = 2, f = f, gradient = f_prime, max = T)$iters, 
                 gradient_desc(starting = 3, f = f, gradient = f_prime, max = T)$iters),
)

comparison_tibble <- g_min_results |>
  inner_join(g_max_results, by = "start_val", suffix = c("_gmin", "_gmax")) |>
  inner_join(n_min_results, by = "start_val") |> 
  rename(x_nmin = x, function_val_nmin = function_val, iterations_nmin = iterations) |>
  inner_join(n_max_results, by = "start_val") |> 
  rename(x_nmax = x, function_val_nmax = function_val, iterations_nmax = iterations)

print(comparison_tibble)

```

It appears that newton's method did better with identifying the max of
this function as only took 12 iterations to converge compared to 22
iterations when both algorithms started at 1. However, it did worse at
identifing the min of this function as it took 58 iterations to converge
compared to 31 iterations when both algorithms started at -1.

### Part g (4 points)

Use the `optimize()` function in **R** to find the global max and min
for the function $f$.

```{r}
#| label: built in optimize

f <- function(x) {
  log(1+x^2)*(1+x)*exp(-x^2)
}

optimize(f, c(-3,3))
optimize(f, c(-3,3), maximum = TRUE)

```

# Problem 2 (26 points)

Suppose $\boldsymbol{x} = (x_1,\dots, x_n )^T$ are independent and
identically distributed random variables with probability density
function (pdf) given by $$
f(x_i|\theta) = \theta(1-x_i)^{\theta-1};\ \ 0\le x_i< 1,\ 0<\theta<\infty,\ i=1,\dots,n
$$

### Part a (4 points)

Using ggplot, plot the probability density function (pdf) for an
individual $x_i$ given $\theta=0.5$ and then again for $\theta=5$.

```{r}
#| label: probability density plot

f <- function(x, theta = theta) {
  theta*(1-x)^(theta-1)
}

ggplot(data.frame(x = c(0, 0.98)), aes(x = x)) +
  facet_wrap(~theta, labeller = label_both) +
  stat_function(data = data.frame(x = c(0, 0.9999), theta = 0.5), # range not to 1 due to dividing by 0 
                fun = f, args = list(theta = 0.5), color = "steelblue") +
  stat_function(data = data.frame(x = c(0, 1), theta = 5), 
                fun = f, args = list(theta = 2), color = "steelblue") +
  labs(y = "Density", x = "x") +
  theme_minimal()

```

### Part b (5 points)

Give the likelihood $L(\theta|\boldsymbol{x})$ and log-likelihood
$\ell(\theta|\boldsymbol{x})$ functions in this situation.

Likelihood Function

$$ 
L(\theta|\boldsymbol{x}) = \prod_{i=1}^{n}\theta(1-x_i)^{\theta-1} \\
L(\theta|\boldsymbol{x}) = \theta^{n}\prod_{i=1}^{n}(1-x_i)^{\theta-1}
$$

Log-Likelihood Function

$$
\ell(\theta|\boldsymbol{x}) = ln(\theta^{n}\prod_{i=1}^{n}(1-x_i)^{\theta-1}) \\
$$

$$
\ell(\theta|\boldsymbol{x}) = ln(\theta^{n}) + ln(\prod_{i=1}^{n}(1-x_i)^{\theta-1}) \\
$$

$$
\ell(\theta|\boldsymbol{x}) = ln(\theta^{n}) + \sum_{i=1}^{n}ln((1-x_i)^{\theta-1}) \\
$$

$$
\ell(\theta|\boldsymbol{x}) = nln(\theta) + \sum_{i=1}^{n}(\theta-1)ln(1-x_i) \\
$$

$$
\ell(\theta|\boldsymbol{x}) = nln(\theta) + (\theta-1)\sum_{i=1}^{n}ln(1-x_i) \\
$$

### Part c (5 points)

Find the Maximum Likelihood Estimator (MLE) $\hat{\theta}$ for $\theta$.

$$
\frac{d\ell}{d\theta} = \frac{n}{\theta} + \sum_{i=1}^{n}ln(1-x_i) = 0 \\
\frac{n}{\theta} = -\sum_{i=1}^{n}ln(1-x_i) \\
\hat{\theta} = -\frac{n}{\sum_{i=1}^{n}ln(1-x_i)}
$$

### Part d (3 points)

Show that your estimator is in fact a maximum. That is, check the second
derivative of the log-likelihood.

$$
\frac{d^{2}\ell}{d\theta^{2}} = -\frac{n}{\theta^{2}} 
$$ 

Since $n$ and $\theta^2$ are both positive,
$\frac{d^{2}\ell}{d\theta^{2}} < 0$ for all $\theta \in (0, \infty)$.
Because the log-likelihood is strictly concave over its entire domain,
the unique critical point $\hat{\theta}$ is the global maximum.

### Part e (2 points)

Find the MLE if the data values are given below. Note, the actual
$\theta$ value used to generate these data was $\theta=7.3$.

```{r}
#| label: mle_values
x <- c(0.0194, 0.0053, 0.2488, 0.0456, 0.2059, 0.0992, 0.1168, 0.3705, 
       0.2129, 0.018, 0.0464, 0.1401, 0.0759, 0.1588, 0.0334, 0.2931, 
       0.0662, 0.2292, 0.1581, 0.3462, 0.035, 0.1086, 0.0793, 0.2095, 
       0.1419, 0.1835, 0.1107, 0.0764, 0.1331, 0.042, 0.0911, 0.1608)

theta_hat <- function(x) {
  n <- length(x)
  running_sum_log <- 0
  for (i in x) {
    current_log <- log(1 - i)
    running_sum_log <- running_sum_log + current_log}
  estimate_theta <- -n / running_sum_log
  return(estimate_theta)
}

theta_hat(x)

```

### Part f (7 points)

To get an idea of how variable the maximum likelihood estimator is in
this case, let's generate many samples of size 32 from this distribution
(which is a Beta distribution with parameters $1$ and $\theta$, but you
don't really need to know that). We can do this by using
`rbeta(32, 1, 7.3)`.

Using the code below, generate at least 10,000 samples this way, compute
the MLE for each, and then plot a histogram of the values using ggplot.
Plot a vertical line at the true $\theta$ value of 7.3 (you can use
`geom_vline(xintercept = 7.3)` for this). On the plot somewhere,
indicate what the standard deviation in those estimates is. Of course,
this code is incomplete, so you'll have to complete it. Once you do,
make sure to change the `eval` option to `true`.

Once you've completed the code for the samples of size 32, repeat this
but instead of taking samples of size 32, take samples of size 100. How
does this change the histogram, where the distribution is centered, and
the standard deviation of the estimates, if at all?

```{r}
#| label: mle_sim_32
#| eval: false # Change this to true once you fill in the code
#| fig-width: 6
#| fig-height: 5
#| out-width: "70%"
#| warning: false
set.seed(2026)
theta_hat <- rep(0, 10000) # Initialize the vector
for(i in 1:10000) {
  x_sim <- rbeta(32, 1, 7.3)
  n <- 32
  running_sum_log <- 0 
  for (j in x_sim) {
    current_log <- log(1 - j)
    running_sum_log <- running_sum_log + current_log
  }
  theta_hat[i] <- -n / running_sum_log
}

ggplot(data.frame(theta_hat), aes(x = theta_hat)) +
  geom_histogram(fill = "steelblue", color = "white") +
  theme_minimal()

```

# Problem 3 (31 points)

We implemented optimization to maximize the likelihood for a Poisson
regression problem in the notes to estimate the $\beta$ (intercept and
slope) values. Let's do something similar for find the
$\boldsymbol{\beta}$ vector by maximizing the likelihood in logistic
regression.

In logistic regression, we attempt to predict the probability of a
success for a given case. We can use the Bernoulli random variable for
this. For a Bernoulli random variable,
$P(Y=y_i) = p_i^{y_i}(1-p_i)^{1-y_i}$ for $y_i=0,1$ where $p_i$ is the
probability of a success. In our case, for logistic regression, we have
$\text{logit}(p_i)=\ln\left(\dfrac{p_i}{1-p_i}\right)=\mathbf{X}_i\boldsymbol{\beta}$.
That means
$p_i = \dfrac{\exp(\mathbf{X}_i\boldsymbol{\beta})}{1+\exp(\mathbf{X}_i\boldsymbol{\beta})}$.

The likelihood function,
$L(\boldsymbol{\beta}|\mathbf{X},\boldsymbol{y})$, for the logistic
regression for a sample of size $n$ is $$
\begin{align*}
L(\boldsymbol{\beta}|\mathbf{X},\boldsymbol{y}) &=\prod_{i=1}^n p_i^{y_i}(1-p_i)^{1-y_i}\\
&=\prod_{i=1}^n\left[\dfrac{\exp(\mathbf{X}_i\boldsymbol{\beta})}{1+\exp(\mathbf{X}_i\boldsymbol{\beta})}\right]^{y_i}\left(1-\left[\dfrac{\exp(\mathbf{X}_i\boldsymbol{\beta})}{1+\exp(\mathbf{X}_i\boldsymbol{\beta})}\right]\right)^{1-y_i}
\end{align*}
$$

### Part a (7 points)

Using the above likelihood, show that the log likelihood function is 
$$
\ell(\boldsymbol{\beta}|\mathbf{X},\boldsymbol{y})=\sum_{i=1}^ny_i\ln\left(\dfrac{\exp(\mathbf{X}_i\boldsymbol{\beta})}{1+\exp(\mathbf{X}_i\boldsymbol{\beta})}\right) + \sum_{i=1}^n(1-y_i)\ln\left(\dfrac{1}{1+\exp(\mathbf{X}_i\boldsymbol{\beta})}\right)
$$ 
and then show it can be equivalently written 
$$
\ell(\boldsymbol{\beta}|\mathbf{X},\boldsymbol{y}) = \sum_{i=1}^ny_i(\mathbf{X}_i\boldsymbol{\beta}) - \sum_{i=1}^n\ln\Big(1 + \exp(\mathbf{X}_i\boldsymbol{\beta})\Big)
$$ 
using properties of logarithms.

$$
L(\boldsymbol{\beta}|\mathbf{X},\boldsymbol{y}) = \prod_{i=1}^n\left[\dfrac{\exp(\mathbf{X}_i\boldsymbol{\beta})}{1+\exp(\mathbf{X}_i\boldsymbol{\beta})}\right]^{y_i}\left(1-\left[\dfrac{\exp(\mathbf{X}_i\boldsymbol{\beta})}{1+\exp(\mathbf{X}_i\boldsymbol{\beta})}\right]\right)^{1-y_i} \\
$$

$$
\ell(\boldsymbol{\beta}|\mathbf{X},\boldsymbol{y}) = \ln{\prod_{i=1}^n\left[\dfrac{\exp(\mathbf{X}_i\boldsymbol{\beta})}{1+\exp(\mathbf{X}_i\boldsymbol{\beta})}\right]^{y_i}\left(1-\left[\dfrac{\exp(\mathbf{X}_i\boldsymbol{\beta})}{1+\exp(\mathbf{X}_i\boldsymbol{\beta})}\right]\right)^{1-y_i}}
$$

$$
\left(1-\left[\dfrac{\exp(\mathbf{X}_i\boldsymbol{\beta})}{1+\exp(\mathbf{X}_i\boldsymbol{\beta})}\right]\right) = \left(\left[\dfrac{1+\exp(\mathbf{X}_i\boldsymbol{\beta})-\exp(\mathbf{X}_i\boldsymbol{\beta})}{1+\exp(\mathbf{X}_i\boldsymbol{\beta})}\right]\right) =  \left(\left[\dfrac{1}{1+\exp(\mathbf{X}_i\boldsymbol{\beta})}\right]\right)
$$

$$
\ell(\boldsymbol{\beta}|\mathbf{X},\boldsymbol{y}) = \sum_{i=1}^n\ln{\left[\dfrac{\exp(\mathbf{X}_i\boldsymbol{\beta})}{1+\exp(\mathbf{X}_i\boldsymbol{\beta})}\right]^{y_i}\left(\left[\dfrac{1}{1+\exp(\mathbf{X}_i\boldsymbol{\beta})}\right]\right)^{1-y_i}} \\
$$

$$
\ell(\boldsymbol{\beta}|\mathbf{X},\boldsymbol{y}) = \sum_{i=1}^n\ln{\left[\dfrac{\exp(\mathbf{X}_i\boldsymbol{\beta})}{1+\exp(\mathbf{X}_i\boldsymbol{\beta})}\right]^{y_i}}+\sum_{i=1}^n\ln{\left(\left[\dfrac{1}{1+\exp(\mathbf{X}_i\boldsymbol{\beta})}\right]\right)^{1-y_i}} \\
$$

$$
\ell(\boldsymbol{\beta}|\mathbf{X},\boldsymbol{y}) = \sum_{i=1}^n{y_i}\ln{\left[\dfrac{\exp(\mathbf{X}_i\boldsymbol{\beta})}{1+\exp(\mathbf{X}_i\boldsymbol{\beta})}\right]}+\sum_{i=1}^n(1-y_i)\ln{\left[\dfrac{1}{1+\exp(\mathbf{X}_i\boldsymbol{\beta})}\right]}
$$

$$
\ell(\boldsymbol{\beta}|\mathbf{X},\boldsymbol{y}) = 
\sum_{i=1}^ny_i\ln\left[\exp(\mathbf{X}_i\boldsymbol{\beta})\right] - y_i\ln\left[1+\exp(\mathbf{X}_i\boldsymbol{\beta})\right] + \ln\left[1\right] - \ln\left[1+\exp(\mathbf{X}_i\boldsymbol{\beta})\right] - (y_i\ln\left[1\right] - y_i\ln\left[1+\exp(\mathbf{X}_i\boldsymbol{\beta})\right])\\
$$

$$
\ell(\boldsymbol{\beta}|\mathbf{X},\boldsymbol{y}) = 
\sum_{i=1}^ny_i\mathbf{X}_i\boldsymbol{\beta} - y_i\ln\left[1+\exp(\mathbf{X}_i\boldsymbol{\beta})\right] + 0 - \ln\left[1+\exp(\mathbf{X}_i\boldsymbol{\beta})\right] - (0 - y_i\ln\left[1+\exp(\mathbf{X}_i\boldsymbol{\beta})\right])) \\
$$

$$
\ell(\boldsymbol{\beta}|\mathbf{X},\boldsymbol{y}) = 
\sum_{i=1}^ny_i\mathbf{X}_i\boldsymbol{\beta} - y_i\ln\left[1+\exp(\mathbf{X}_i\boldsymbol{\beta})\right] - \ln\left[1+\exp(\mathbf{X}_i\boldsymbol{\beta})\right] + y_i\ln\left[1+\exp(\mathbf{X}_i\boldsymbol{\beta})\right]\\
$$

$$
\ell(\boldsymbol{\beta}|\mathbf{X},\boldsymbol{y}) = 
\sum_{i=1}^ny_i\mathbf{X}_i\boldsymbol{\beta} - \ln\left[1+\exp(\mathbf{X}_i\boldsymbol{\beta})\right] 
$$

$$
\ell(\boldsymbol{\beta}|\mathbf{X},\boldsymbol{y}) = 
\sum_{i=1}^ny_i\mathbf{X}_i\boldsymbol{\beta} - \sum_{i=1}^n\ln\left[1+\exp(\mathbf{X}_i\boldsymbol{\beta})\right] \\
$$

### Part b (6 points)

Find the gradient of the log likelihood if we have three $\beta$'s. That
is, if
$\mathbf{X}_i\boldsymbol{\beta}=\beta_0+\beta_1x_{i1}+\beta_2x_{i2}$.

$$
\frac{d\ell}{d\boldsymbol{\beta}_0} = 
\sum_{i=1}^n\left[y_i - \frac{\exp(\boldsymbol{\beta}_0 + \boldsymbol{\beta}_1\mathbf{X}_{i1} + \boldsymbol{\beta}_2\mathbf{X}_{i2})}{1+\exp(\boldsymbol{\beta}_0 + \boldsymbol{\beta}_1\mathbf{X}_{i1} + \boldsymbol{\beta}_2\mathbf{X}_{i2})}\right] = 
\sum_{i=1}^n\left[y_i-p_i\right] \\
$$

$$
\frac{d\ell}{d\boldsymbol{\beta}_1} = 
\sum_{i=1}^n\left[y_i\mathbf{X}_{i1} -\frac{\exp(\boldsymbol{\beta}_0 + \boldsymbol{\beta}_1\mathbf{X}_{i1} + \boldsymbol{\beta}_2\mathbf{X}_{i2})}{1+\exp(\boldsymbol{\beta}_0 + \boldsymbol{\beta}_1\mathbf{X}_{i1} + \boldsymbol{\beta}_2\mathbf{X}_{i2})}\right] = \sum_{i=1`}^n\left[y_i\mathbf{X}_{i1} - p_i\mathbf{X}_{i1}\right] = \sum_{i=1}^n\left[\mathbf{X}_{i1}(y_i-p_i)\right] \\
$$

$$
\frac{d\ell}{d\boldsymbol{\beta}_2} = 
\sum_{i=1}^n\left[y_i\mathbf{X}_{i2} -\frac{\exp(\boldsymbol{\beta}_0 + \boldsymbol{\beta}_1\mathbf{X}_{i1} + \boldsymbol{\beta}_2\mathbf{X}_{i2})}{1+\exp(\boldsymbol{\beta}_0 + \boldsymbol{\beta}_1\mathbf{X}_{i1} + \boldsymbol{\beta}_2\mathbf{X}_{i2})}\right] = \sum_{i=1`}^n\left[y_i\mathbf{X}_{i2} - p_i\mathbf{X}_{i2}\right] = \sum_{i=1}^n\left[\mathbf{X}_{i2}(y_i-p_i)\right]
$$

### Part c (4 points)

In Homework 3, Problem 2, part a, you read in the `adult` dataset (from
the UC Irvine [database](https://archive.ics.uci.edu/dataset/2/adult)).
This is the one that we used to predict whether a person makes over
\$50K a year based on some other variables. The data came from the
Census Bureau in 1994 and can be found in the Data folder in my
Math3190_S26 GitHub repo. More info on the dataset can be found in the
"adult.names" file.

Read in the dataset and put column names like you did in HW 3. Then fit
a logistic regression model using the `glm()` function that predicts
salary from age/10 and sex. Note, we should divide the age by 10 in our
model because this makes the gradient descent more stable (for whatever
reason). We can divide by 10 in the `glm()` function by wrapping it in
the `I()` function. Like this:
`glm(salary ~ I(age/10) + sex, data = adult, family = binomial)`.

```{r}
#| label: read in data

adult_names <- read_csv("/Users/melaniejensen/Library/CloudStorage/OneDrive-SouthernUtahUniversity/Math_3190/Math_3190_Assignments/Homework/Homework 3/adult.names") |> 
  tail(15)

adults <- read_csv("/Users/melaniejensen/Library/CloudStorage/OneDrive-SouthernUtahUniversity/Math_3190/Math_3190_Assignments/Homework/Homework 3/adult.data") |> 
  rename(age = '39', work_class = 'State-gov', fnlwgt = '77516', education = 'Bachelors', education_num = '13', marital_status = 'Never-married', occupation = 'Adm-clerical', relationship = 'Not-in-family', race = 'White', sex = 'Male', capital_gain = '2174', capital_loss = '0', hours_per_week = '40', native_country = 'United-States', salary = '<=50K') |> 
  mutate(salary = factor(salary, levels = c('<=50K', '>50K')))

```

```{r}
#| label: logistic regression

mod1 <- glm(salary ~ I(age/10) + sex, data = adults, family = binomial)
summary(mod1)

```

### Part d (4 points)

Define `y`, `x1`, and `x2`. `y` should be a vector of 0's and 1's with a
1 indicating that the person had a salary above \$50,000. `x1` should be
a vector that contains all of the age values (divided by 10) and then
`x2` should be an indicator variable that indicates if the person is
male or not. Taking columns from the matrix obtained using the
`model.matrix()` function applied to your logistic regression model you
fit in part c may be useful here.

```{r}
#| label: Definitions

X_matrix <- model.matrix(mod1)
x1 <- X_matrix[, "I(age/10)"]
x2 <- X_matrix[, "sexMale"]
y <- as.numeric(adults$salary) - 1

```

### Part e (7 points)

Use the gradient descent function from problem 1 to find estimates for
$\beta_0$, $\beta_1$, and $\beta_2$ for predicting salary from age and
sex. Enter the following in your function:

-   Use the vector `c(0, 0, 0)` as your starting point.
-   Instead of starting the step size, $\gamma_k$, at 1, start it at 0.1
    to save some time with the line search.
-   Make sure the maximum number of iterations is at least 200.

You'll know you did this right if the optimized values you end up with
match (or are very close to) the estimates from your logistic regression
model you fit in part c. This may take a minute or two to run, so you
may want to cache this code chunk.

```{r}
#| label: logistic gradient descent

f <- function(beta, y, x_long) {
  n <- length(y)
  X <- cbind(1, x_long[1:n], x_long[(n+1):(2*n)])
  z <- X %*% beta
  val <- sum(y * z - log(1 + exp(z)))
  
  return(as.vector(val))
}

f_prime <- function(beta, y, x_long) {
  n <- length(y)
  X <- cbind(1, x_long[1:n], x_long[(n+1):(2*n)])
  z <- X %*% beta
  p <- exp(z) / (1 + exp(z))
  grad <- t(X) %*% (y - p)
  return(as.vector(grad))
}

gradient_desc(starting = c(0, 0, 0), f = function(b) f(b, y, c(x1,x2)), 
  gradient = function(b) f_prime(b, y, c(x1,x2)), max = TRUE, maxit = 500, 
  step_size = 0.1)

```

### Part f (3 points)

Use the `optim()` function in **R** to find the estimate for the
$\beta$'s here. This should be much faster since this function is much
better optimized (no pun intended). The results here should also be very
close to the numbers we obtained for the slopes in the model we fit in
part c, but may not match exactly due to different stopping criteria.

```{r}
#| label: built in optim()

optim(par = c(0, 0, 0),fn = function(b) -f(b, y, c(x1,x2)),
  gr = function(b) -f_prime(b, y, c(x1,x2)))

```


$\[\begin{align*} f'(x) &= \frac{d}{dx}\left[\ln(1+x^2)\right](1+x)e^{-x^2} + \ln(1+x^2)\frac{d}{dx}\left[(1+x)e^{-x^2}\right]\\ &= \frac{d}{dx}\left[\ln(1+x^2)\right](1+x)e^{-x^2} + \ln(1+x^2)\left(\frac{d}{dx}\left[(1+x)\right]e^{-x^2} + (1+x)\frac{d}{dx}\left[e^{-x^2}\right]\right)\\ &= \dfrac{2x}{1+x^2}(1+x)e^{-x^2} + \ln(1+x^2)\left(1\cdot e^{-x^2} + (1+x)(-2x)e^{-x^2}\right)\\ &= e^{-x^2}\left(\dfrac{2x(1+x)}{1+x^2} + \ln(1+x^2) - 2x(1+x)\ln(1+x^2)\right)\\ &=e^{-x^2}\left(\dfrac{2x(1+x)}{1+x^2} + \ln(1+x^2)\big(1 - 2x-2x^2\big)\right) \end{align*}\]$