---
title: "**MATH 3190 Homework 7**"
subtitle: |
  Due March 26, 2026\
  Focus: Notes 8\
  Total Points: 90 + 10 (Self-Critique Comment and Uploading to GitHub) = 100
format: 
  html:
    embed-resources: true # Keep this so the html file is easily shareable.
urlcolor: blue
execute: # Will set these as the default for all code chunks
  eval: true
  echo: true
  cache: true
editor_options: 
  chunk_output_type: console
---

Note that your homework should be completed in Quarto. You can just add your answers and code to this document in the appropriate part. The file should be rendered as an html document or pdf document (html is quicker to render and you will need an installation of LaTeX to render as a pdf). You will "turn in" this homework by uploading the files, the .qmd and output file, to your GitHub `Math_3190_Assignment` repository in the `Homework/Homework_7` directory. Make sure all supporting files (if there are any) are in this directory as well. You should use `git` and `Unix` commands in a terminal for this (i.e. don't manually upload files to GitHub). In addition, you will submit just your html or pdf file on Canvas to unlock the solutions.

Don't forget to also put a [self-critique comment]{.underline} on Canvas indicating what you did correctly and incorrectly and include any questions you may have about the assignment after submitting the assignment and thoroughly looking through the solutions.

```{r}
#| label: initializations
#| hidden: true

library(tidyverse)
library(rsample)
library(GGally)
library(car)
library(pls)
library(umap)
library(math3190package)

```

# Problem 1 (57 points)

In Homework 6, we briefly looked at a small sample from this data set. Now we have the full data set with 20,640 districts. For districts in California from 1990, we want to predict the median house value for the district based on the district location (longitude and latitude), the median house age in the block group, the total number of rooms in the homes, the total number of bedrooms, the population in the district, the number of households, and the median income (in $10,000). These data were obtained from [scikit-learn.org](https://scikit-learn.org/stable/datasets.html) and can be found in the Data folder of the Math3190_Sp26 GitHub repo with the name `cal_housing.data`. 

### Part a (5 points)
Read in the data. The `read.csv()` function works for files that end in `.data` if they are comma separated like this file is. However, be aware that the housing data does not have column names, so make sure to put `header = FALSE`. Assign appropriate column names to the columns and then take logs (base e) of all variables except lat and long. Once this is done, split the data into training and testing sets with the testing containing 20% of the values. Set a seed when you do this.

```{r}
#| label: p1a

cal_housing <- read.csv('/Users/melaniejensen/Library/CloudStorage/OneDrive-SouthernUtahUniversity/Math_3190/Math_3190_Assignments/Homework/Homework 7/cal_housing.data', header = FALSE) |> 
  setNames(c('longitude', 'latitude', 'housingMedianAge', 'totalRooms', 'totalBedrooms', 'population', 'households', 'medianIncome', 'medianHouseValue')) |> 
  mutate(across(-c(longitude, latitude), log))

set.seed(2026)
split <- initial_split(cal_housing, prop = 0.8)
housing_train <- training(split) |> as.tibble()
housing_test  <- testing(split) |> as.tibble()

```

### Part b (5 points)
Use the following command to view pair-wise scatter plots for the data here. Change `housing_train` to whatever you called the training set and be sure to change the `eval` option for the code chunk to `true`. Note, that `lower` option changes the points to be plotted with periods instead of circles. This speeds up plotting time considerably and makes the plots more readable since there are many data points here.

Based on these plots, does multicollinearity appear to be an issue here? Explain.

```{r}
#| label: p1b
#| cache: true
#| eval: true

ggpairs(housing_train, 
        lower = list(continuous = wrap("points", shape = "."))
)

```

Based on these plots, it does appear that multicollinearity will likely be an issue because it appears that totalRooms, totalBedrooms and population are all highly correlated with one another.

### Part c (4 points)
Fit a OLS regression model predicting house prices from all other variables. Check its summary output and its VIF values. Comment on the VIFs.

```{r}
#| label: p1c

linear_housing_mod <- lm(medianHouseValue ~ ., housing_train)
summary(linear_housing_mod)
vif(linear_housing_mod)

```

After looking at the VIF values, we can see that multicollinearity is a big issue in this data, as 3 variables have a VIF score well above 10, and an additional 3 variables have a VIF score close to 10.

### Part d (3 points)
Fit a principal component model using the `pcr()` function in the `pls` package. Since the variables are on very different scales, use the `scale = TRUE` option in the function. Then find the VIF values for this model. The `vif()` function from the `car` library won't work here. Instead, you can take the diagonals of the inverse of the correlation matrix (like is shown in the regularization section of Notes 7) for the `scores` output. The code for this is `diag(solve(cor(pca_model$scores)))`. Of course, you need to change `pca_model` to what you called your model.

```{r}
#| label: p1d

pcr_housing_mod <- pcr(medianHouseValue ~ ., data = housing_train, scale = TRUE)
summary(pcr_housing_mod)
diag(solve(cor(pcr_housing_mod$scores)))

```

### Part e (5 points)
Now take the summary of your principal component model. With PCA, the common amount of variation we want to explain in the predictors is 90% or more. How many components are needed to achieve this? For that number of components, how much of the variation in log of home values is explained? How many components are needed to explain a "good" amount of the variation in the log of home values? Note: the upper bound on the amount of variation in the log of home values here with all 8 components will be equal to the $R^2$ value of the OLS model. 

```{r}
#| label: p1e

summary(pcr_housing_mod)

```

To have 90% or more of the variation in predictors explained, 4 components are required. For having 4 components, 50.16% of the variation in the log home values is explained.
I would say that having 4 components explains a 'good' amount of the variation in logged home values because the marginal increase in variation in logged home value is much smaller going from 4-5 components than it was for going between 3-4 componenets. 

### Part f (4 points)
Now fit a partial least squares model using the `plsr()` function in the `pls` package with `scale = TRUE`. Then find the VIF values for this model like you did in part (d) and comment on them.

```{r}
#| label: p1f

plsr_housing_mod <- plsr(medianHouseValue ~ ., data = housing_train, scale = TRUE)
summary(plsr_housing_mod)
diag(solve(cor(plsr_housing_mod$scores)))

```

The VIF scores are all coming out to be 1, which makes sense because each component is orthogonal to the other.

### Part g (4 points)
Now take the summary of your partial least squares model. Compare this output to the summary of the PCA model. Explain what the differences are. 

How many components are needed to balance a good amount of variation explained in $X$ and in $y$ for the PLS model?

```{r}
#| label: p1g

summary(pcr_housing_mod)
summary(plsr_housing_mod)

```

Comparing the two models, the PLSR model explains variation in log median house value more efficiently — reaching near its maximum R² of 67.57% in fewer components than the PCR model. This is expected, as PLSR constructs components by maximizing covariance with the outcome, whereas PCR constructs components solely based on variation in the predictors without regard to the outcome.

Conversely, the PCR model captures variation in the predictors more efficiently in the first few components, where 4 components in the PLSR explains ~91% compared to the PCR model's explained ~97% of predictor variance. This tradeoff reflects the fundamental difference between the two methods — PCR prioritizes explaining the predictors while PLSR prioritizes explaining the outcome.

In this model, a 'good' number of components is 5 components as the difference in marginal variation in loged median house value is smaller between 5-6 components than it was 4-5. 

### Part h (8 points)
In part (e), you said how many components were needed for the PCA to explain a good amount of the $Y$ variable. Using that number of components, use the PCA and PLS models to find the root MSE when predicting the testing set. Note: the root MSE (or residual standard error, also abbreviated RMSE) is found by taking the square root of the sum of squared residuals divided by the residual degrees of freedom, which is $n-p-1$, where $p$ is the number of predictors in the model. Also, be aware that when you use the `predict()` function on your model, it will give you predictions for each number of components. You can access the predictions for using 3 components, for example, by typing `predict(pca_model, test_set)[,,3]`.

Then in part (g), you said how many components were needed for the PLS Using that number of components. Using this number of components, use the PCA and PLS models to find the root MSE predicting the testing set.

Compare the results of the predictions in both cases. 

```{r}
#| label: p1h1

pcr_pred <- predict(pcr_housing_mod, newdata = housing_test, ncomp = 4)
pcr_rmse <- sqrt(mean((housing_test$medianHouseValue - pcr_pred)^2))

plsr_pred <- predict(plsr_housing_mod, newdata = housing_test, ncomp = 4)
plsr_rmse <- sqrt(mean((housing_test$medianHouseValue - plsr_pred)^2))

pcr_rmse
plsr_rmse

```

The PLSR model outperformed the PCR model in terms of predictive accuracy, achieving a lower test RMSE of 0.362 compared to 0.409 for PCR (in log units). This is consistent with the summary output, as PLSR explained a greater proportion of variation in log median house value with fewer components, reflecting its advantage of directly optimizing for the outcome variable during component construction.

```{r}
#| label: p1h2

pcr_pred_new <- predict(pcr_housing_mod, newdata = housing_test, ncomp = 5)
pcr_rmse_new <- sqrt(mean((housing_test$medianHouseValue - pcr_pred_new)^2))

plsr_pred_new <- predict(plsr_housing_mod, newdata = housing_test, ncomp = 5)
plsr_rmse_new <- sqrt(mean((housing_test$medianHouseValue - plsr_pred_new)^2))

pcr_rmse_new
plsr_rmse_new

```

Increasing to 5 components had a negligible effect on PCR predictive accuracy but meaningfully improved PLSR, suggesting the optimal number of components differs between the two models. PLSR with 5 components achieved the lowest test RMSE of 0.329, making it the best performing model overall.

### Part i (12 points)
Using the number of components you said were needed for the PLS model in part (g), come up with reasonable surrogates for each of those components. Look at the projection output for this. When creating your surrogates, make sure to center and scale each variable. You can do this with the `scale()` function and putting your housing data frame or tibble in it.

Then fit a OLS model using those surrogates in the training set. Check the VIF of this model to make sure each one is below 5. If they are not, your surrogates should be changed. You can use the `vif()` function in the `car` package here.

```{r}
#| label: p1i1

loadings(plsr_housing_mod)[, 1:5]

scaled_housing_train <- scale(housing_train) |> as.data.frame()
scaled_housing_test <- scale(housing_test, 
                       center = attr(scale(housing_train), "scaled:center"), 
                       scale = attr(scale(housing_train), "scaled:scale")) |> as.data.frame()

housing_list <- list(train = scaled_housing_train, test = scaled_housing_test)

for (n in seq_along(housing_list)) {
  surrogate2_calc_vars <- c("totalRooms", "totalBedrooms", "households", "population")
 
  housing_list[[n]]$comp1_surrogate <- housing_list[[n]]$medianIncome
  housing_list[[n]]$comp2_surrogate <- rowMeans(housing_list[[n]][, surrogate2_calc_vars])
  housing_list[[n]]$comp3_surrogate <- housing_list[[n]]$longitude / housing_list[[n]]$households
  housing_list[[n]]$comp4_surrogate <- housing_list[[n]]$latitude
  housing_list[[n]]$comp5_surrogate <- housing_list[[n]]$housingMedianAge
}

scaled_housing_train <- housing_list$train
scaled_housing_test <- housing_list$test

```

```{r}
#| label: p1i2

linear_surragate_housing_mod <- lm(medianHouseValue ~ comp1_surrogate + comp2_surrogate + comp3_surrogate + comp4_surrogate + comp5_surrogate, scaled_housing_train)
summary(linear_surragate_housing_mod)

vif(linear_surragate_housing_mod)

```

### Part j (7 points)
Now using that model you fit in the previous part with the surrogates, find the root MSE for predicting the testing set using that model, and compare it to what you got in part (h). Describe some pros and cons of the surrogate model.

Hint: to predict using this model, you will have to create a `newdata` data frame (or tibble) and then redefine your surrogates in that data frame using the scaled testing data set. Make sure the variable names in the `newdata` data frame match the variables names used in the model you defined in part (i). 

```{r}
#| label: p1j

surrogate_cols <- c("comp1_surrogate", "comp2_surrogate", "comp3_surrogate", 
                    "comp4_surrogate", "comp5_surrogate")

surrogate_preds <- predict(linear_surragate_housing_mod, 
                           newdata = scaled_housing_test[, surrogate_cols])

surrogate_rmse <- sqrt(mean((scaled_housing_test$medianHouseValue - surrogate_preds)^2))
surrogate_rmse

```

# Problem 2 (33 points)

In base **R**, there is a dataset called `penguins`. Type `?penguins` to read a little about the data.

### Part a (6 points)
Remove the rows in the penguins dataset that have missing values. Then use the `ggpairs()` function from `GGally` to make pairwise scatterplots for the four quantitative variables and color the points based on the `species` variable. Note that the `ggpairs()` function has a `columns` argument where you can select the columns you want to plot and it has a `mapping` argument where you can put `aes(color = species, alpha = 0.5)` to color the points by species. 

Include the plot here and then comment on which scatterplots seem to nicely separate the penguins by species and which tend to have worse separation. 

```{r}
#| label: p2a

penguins <- penguins |> 
  na.omit()

ggpairs(penguins, aes(color = species, alpha = 0.5), upper = list(continuous = wrap("cor", size = 2))) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 3),
        axis.text.y = element_text(hjust = 1, size = 3))

```

After looking at the scatter plots, it appears that bill_len combination with bill_dep, flipper_len, and body_mass all seperate the three species very well into clear clusterings. 

### Part b (4 points)
Run a scaled principal component analysis on the four quantitative variables using the `prcomp()` function. How much of the total variation in these variables is explained by the first two principal components? You can use the `summary()` function on your PCA to find this easily. 

```{r}
#| label: p2b

qualitative_penguins <- penguins |> 
  select(bill_len, bill_dep, flipper_len, body_mass)

pca <- prcomp(qualitative_penguins, scale = T) 
summary(pca)

```

88.09% of the variation in the variables is explained by the first two principals as shown in the cumulative proportion.

### Part c (6 points)
Plot the first two principal components with PC1 on the x-axis and PC2 on the y-axis. Color the points by species and comment on how well the first two principal components are separating by species. 

```{r}
#| label: p2c

pca_df <- pca$x |> 
  as_tibble() |> 
  mutate(species = penguins$species)

ggplot(pca_df, aes(x = PC1, y = PC2, color = penguins$species)) + 
  geom_point()

```

The first two components do a good job and seperating the Gentoo species from the masses. While there is some seperation between Adelie and Chinstrap, it's not nearly as seperated as how Gentoo is seperated out. 

### Part d (12 points)
Now use the `umap()` function in the `umap` library to fit a UMAP to the **scaled** quantitative variables in the penguins dataset. You can use the `scale()` function to scale the variables. Make sure the UMAP gives two components and set a `random_state` in the function so your results are reproducible. Then plot the UMAP using the the `umap_plot()` function in `math3190package` and color the points by `species` using the `color_umap` argument. Experiment with at least three different number of neighbors (`n_neighbors`) and at least three different minimum distances (`min_dist`).

```{r}
#| label: p2d

neighbors_list <- c(5, 15, 30)
dist_list <- c(0.01, 0.1, 0.5)

for (n in neighbors_list) {
  for (d in dist_list) {
    
    penguin_umap <- umap(scale(qualitative_penguins), n_components = 2, random_state = 2026, 
                         n_neighbors = n, min_dist = d)
    
    p <- umap_plot(penguin_umap, data = penguins, color_umap = "species", size = 2) + 
      scale_colour_viridis_d() + 
      labs(title = paste0("UMAP Plot (n_neighbors = ", n, ", min_dist = ", d, ")"))
    
    print(p)
  }
}

```

### Part e (5 points)
Comment on the UMAP plots you made in part (d). Include a comparison to the PCA plot regarding how well these species are separated. 

In creating the umaps, I found that a min_distance of 0.5 seperated the species pretty well, specifically with 5 and 30 neighbors. 

I found that like pca, the gentoo species was seperating very well where the other two species tend to be intertwined. However, those two species seperate better in the umap than they did in the pca. 

Feel free to repeat this problem using a variable other than species! Sometimes the UMAP outperforms PCA at separating classes, sometimes it does a little worse, and sometimes there are both quite bad. 
