---
title: "**MATH 3190 Homework 5**"
subtitle: |
  Due February 26, 2026\
  Focus: Boostrapping from Notes 5 and Notes 7 Part 1\
  Total Points: 90 + 10 (Self-Critique Comment and Uploading to GitHub) = 100
format: 
  html:
    embed-resources: true # Keep this so the html file is easily shareable.
    toc: true
    toc-depth: 3
    toc-location: left
    toc-title: "Page Navigation"
    toc-expand: true
urlcolor: blue
execute: # Will set these as the default for all code chunks
  eval: true
  echo: true
  cache: true
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
engine: knitr
---

Note that your homework should be completed in Quarto. You can just add your answers and code to this document in the appropriate part. The file should be rendered as an html document or pdf document (html is quicker to render and you will need an installation of LaTeX to render as a pdf). You will "turn in" this homework by uploading the files, the .qmd and output file, to your GitHub `Math_3190_Assignment` repository in the `Homework/Homework_5` directory. Make sure all supporting files (if there are any) are in this directory as well. You should use `git` and `Unix` commands in a terminal for this (i.e. don't manually upload files to GitHub). In addition, you will submit just your html or pdf file on Canvas to unlock the solutions.

Don't forget to also put a [self-critique comment]{.underline} on Canvas indicating what you did correctly and incorrectly and include any questions you may have about the assignment after submitting the assignment and thoroughly looking through the solutions.

Also, there is a package I created with many functions and data files we will be using for the remainder of this class. You can install this package instead of downloading and reading in the data files. To download the package, do the following in **R**:

1. Install the remotes or the `devtools` package using either `install.packages("remotes")` or `install.packages("devtools")`.
2. Load the package: `library(remotes)` or `library(devtools)`.
3. Install the course package: `install_github("rbrown53/math3190package")`. If it asks you if you want to update packages, you can select that you do. 
4. Now, you will simply have to load the package (which will also load the `tidyverse`) for each homework assignment: `library(math3190package)`.

```{r}
#| label: setup

library(devtools)
library(remotes)
library(math3190package)

```

## Problem 1 (35 points)

We first covered bootstrapping in Notes 5, but we returned to it here. Let's do a problem on bootstrapping that takes ideas from Notes 5 and Notes 7. 

The amount of money spent on health care is an important issue for workers because many companies provide health insurance that only partial covers many medical procedures. The director of employee benefits at a midsize company wants to determine the amount spent on health care by the typical hourly worker in the company. A random sample of 25 workers is selected and the amounts they spent on their family’s health care needs during the past year can be found in the `healthcare` tibble in the `math3190package`. 

```{r}
#| label: healthcare_read-in

healthcare <- healthcare

```

### Part a (4 points)
Graph the data using both a boxplot and normal quantile plot (QQ plot) and use these to assess whether the population has a normal distribution.

```{r}
#| label: healthcare_plots

ggplot(healthcare, aes(x = "", y = amount)) +
  geom_boxplot() +
  labs(title = "Boxplot of Healthcare Costs", y = "Cost")

ggplot(healthcare, aes(sample = amount)) + 
  geom_qq() +
  geom_qq_line() +
  labs(title = "QQ Plot of Healthcare Costs")

```

The healthcare amounts are not normally distributed. The boxplot shows that the data is skewed to the right, and the QQ plot is showing signs of a heavy right tail, which violates the normality assumption. 

### Part b (4 points)
Construct a 95% confidence interval for the mean cost per household using a t-interval. You can use the `t.test()` function for this. Interpret this confidence interval in context of the problem and comment on whether the t-procedure is appropriate to use here. 

```{r}
#| label: healthcare_ci_interval

t.test(healthcare$amount, conf.level = 0.95)

```

The t-test CI indicates that the average amount spend on healthcare is between $163.60 and $982.64. However, the t-procedure is unreliable to use here because the normality assumption is violated, as evidenced by the box plot and QQ plot. Having the skew in such a small sample means that the confidence level is underestimated (the actual coverage is less than 95%) and the interval itself is overestimated (too wide). Both occur simultaneously because the wide interval starts too far to the left. 

### Part c (7 points)
Now let’s use bootstrapping to estimate a 95% interval for the mean.

First, write a loop yourself that will do this. In the loop, you’ll need to sample from your data set with replacement, take the mean of that sample, and then store those means. Have the loop do this 10,000 times. Then you can obtain a 95% interval by using the quantile function in **R**. Specifically, you can type `quantile(means_vector, c(0.025, 0.975))` (if you call your vector `means_vector`) to get a 95% interval based on the percentile method.

```{r}
#| label: healthcare_bootstrap_loop

set.seed(2026)
means_vector <- c() 

for (i in 1:10000) {
  sample_data <- sample(healthcare$amount, replace = TRUE)
  means_vector[i] <- mean(sample_data)
}

quantile(means_vector, c(0.025, 0.975))

```

### Part d (5 points)
Now use the `bootstrapping()` function in the `math3190package` library to obtain a bootstrap object using 10,000 samples. Then use the `boot_ci()` function to obtain the 95% intervals. Report the Percentile interval and BCa interval. Compare your interval from part (c) to the Percentile interval computed here. They should be similar. 

```{r}
#| label: healthcare_bootstrap_function

set.seed(2026)

boot_mean <- function(d, i) {
  return(mean(d[i]))
}
bootstrap <- bootstrapping(data = healthcare$amount, boot_mean, 10000)
boot_ci(bootstrap, conf = 0.95, type = "perc")
boot_ci(bootstrap, conf = 0.95, type = "bca")

tibble(
  method = c("Part_c", "Percentile", "BCa"),
  lower = c(quantile(means_vector, 0.025), 261, 306),
  upper = c(quantile(means_vector, 0.975), 997, 1235),
  width = upper - lower
)

```

We can see that the intervals are all pretty similar. The lower bound ranges from 259 to 306 and the upper bound ranges from 997 to 1235. The percentile interval is closest to the interval found in part c, being a slightly skinnier interval. This means that the part c interval is predicting with 95% confidence that the true mean healthcare cost is between $259 and $1004, while the percentile interval is predicting with 95% confidence that the true mean healthcare cost is between $261 and $997. The BCa interval is a bit wider, predicting with 95% confidence that the true mean healthcare cost is between $306 and $1235.

### Part e (3 points)
Write a sentence or two comparing the intervals you obtained in parts (c) and (d) to the t-interval you constructed in part (b). 

```{r}
#| label: healthcare_interval_comparison

tibble(
  method = c("t-test", "Part_c", "Percentile", "BCa"),
  lower = c(163.5954, quantile(means_vector, 0.025), 261, 306),
  upper = c(982.6446, quantile(means_vector, 0.975), 997, 1235),
  width = upper - lower
)

```

The t-test interval is wider than both the part c and percentile intervals, which confirms that the t-test interval is overestimating the margin of error due to the skew in the data. With the t-test starting much sooner than the bootstrapped intervals, it also confirms the shift in the interval due to bias from the skew. 

### Part g (4 points)
Use the `boostrapping()` function with at least 10,000 samples to obtain a 95% confidence interval based on BCa for the median.

```{r}
#| label: healthcare_bootstrap_median

set.seed(2026)
boot_med <- function(d, i) {
  return(median(d[i]))
}
bootstrap_med <- bootstrapping(data = healthcare$amount, boot_med, 10000)
boot_ci(bootstrap_med, conf = 0.95, type = "bca")

```

### Part h (5 points)
The IQR is a measure of spread commonly used with the median. Construct a 95% confidence interval based on BCa for the IQR using the `bootstrapping()` function with at least 10,000 samples and interpret the interval. Also plot the output of the `bootstrapping()` function and comment on whether the distribution looks close to normal.

```{r}
#| label: healthcare_bootstrap_iqr

set.seed(2026)
boot_iqr <- function(d, i) {
  return(IQR(d[i]))
}
bootstrap_iqr <- bootstrapping(data = healthcare$amount, boot_iqr, 10000)
boot_ci(bootstrap_iqr, conf = 0.95, type = "bca")

iqr_plot_data <- tibble(replicates = as.vector(bootstrap_iqr$t))

ggplot(iqr_plot_data, aes(x = replicates)) + 
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  labs(title = "Bootstrap Distribution of IQR",
       x = "Interquartile Range (IQR)",
       y = "Frequency") +
  theme_minimal()

```

The bootstrapping distribution appears to be approximately normal, so long as you don't look at the concentration of outliers on the right side of the distribution. Due to where the outliers fall, it does not appear to be totally normal, as those will skew the descriptive statistics of the distribution. 

### Part i (3 points)
In all of these bootstraps, what did we have to assume about the population in order for these intervals to be valid?

In order for these intervals to be valid, we must assume that the sample is representative of the population and that the observations are independent. Unlike the t-test, the bootstrap intervals do not follow the assumption that the population is normally distributed, so we do not need to assume normality for the bootstrap intervals to be valid.

## Problem 2 (55 points)

Concrete is the most important material in civil engineering. The concrete compressive strength is an important attribute of the concrete and our goal is to predict the concrete compressive strength (in MPa) from the following variables:

* Cement (in $\text{kg}/\text{m}^3$)
* Blast furnace slag (in $\text{kg}/\text{m}^3$)
* Fly ash (in $\text{kg}/\text{m}^3$)
* Water (in $\text{kg}/\text{m}^3$)
* Superplasticizer (in $\text{kg}/\text{m}^3$)
* Coarse aggregate (in $\text{kg}/\text{m}^3$)
* Fine aggregate (in $\text{kg}/\text{m}^3$)
* Age of concrete (in days)

This dataset came from the [UCI ML Repository](https://archive.ics.uci.edu/dataset/165/concrete+compressive+strength) and can be found on the Math3190_Sp26 GitHub page along with a Readme file and in the `Concrete_Data` tibble in the `math3190package` **R** library. 

### Part a (2 points)
Change the names of the variables in `Concrete_Data` so they are shorter yet still descriptive. Name this something other than `Concrete_Data` so you don't overwrite it. 

```{r}
#| label: column_rename

data <- Concrete_Data |> 
  rename(
    "cement" = "Cement (component 1)(kg in a m^3 mixture)",            
    "blast_furnace_slag" = "Blast Furnace Slag (component 2)(kg in a m^3 mixture)",
    "fly_ash" = "Fly Ash (component 3)(kg in a m^3 mixture)",           
    "water" = "Water  (component 4)(kg in a m^3 mixture)",            
    "superplasticizer" = "Superplasticizer (component 5)(kg in a m^3 mixture)", 
    "course_agg" = "Coarse Aggregate  (component 6)(kg in a m^3 mixture)", 
    "fine_agg" = "Fine Aggregate (component 7)(kg in a m^3 mixture)",    
    "age" = "Age (day)",                                            
    "compress_strength" = "Concrete compressive strength(MPa, megapascals)" 
  )

```

### Part b (5 points)
In the `GGally` library is the function `ggpairs()`, which makes a nice scatterplot matrix in the `ggplot2` style. `GGally` should have already been installed when you installed `math3190package`, so you will just need to load it. Using `ggpairs()`, Create this scatterplot matrix for all of the variables in the dataset (they should all be plotted together in one plot). 

Comment on the scatterplot matrix. Which variables seem to have a (at least somewhat) linear relationship with compressive strength? 

```{r}
#| label: ggpairs_matrix

library(GGally)

ggpairs(data)

```

From the ggpairs matrix, it appears that cement has the strongest linear relationship with compressive strength, with a correlation of 0.498 and the scatterplot showing a positive linear-approximate relationship. Superplasticizer is also showing signs of have a positive linear relationship with compressive strength, with a correlation of 0.366, but showing less of a linear relationship in the scatterplot than what we see with cement. While the correlation for age is also relatively higher, at 0.329, the scatterplot is showing a concave shape, which is not linear. 

### Part c (3 points)
Fit a linear model (with the `lm()` function) predicting compressive strength using all other variables. After you've done that, interpret the slope of the water variable in the context of the problem and answer whether it is considered a significant variable in the model. 

```{r}
#| label: linear_model

mod_lm <- lm(compress_strength ~ ., data = data)
summary(mod_lm)

```

An increase of 1 kg/m^3 of water is associated with a decrease in the predicted compressive strength of 0.150298 MPa, holding all other variables constant. The p-value for the water variable is 0.000194, which is less than the common significance level of 0.05, so we would consider water to be a significant variable in the model.

### Part d (6 points)
Using `ggplot()`, make a QQ plot of the raw residuals (obtained using the `residuals()` function). Include the QQ line as well. Comment on whether the residuals appear to be approximately normal.

```{r}
#| label: qq_plot_raw_residuals

ggplot(data, aes(sample = residuals(mod_lm))) + 
  geom_qq() +
  geom_qq_line() +
  labs(title = "QQ Plot of Raw Residuals") + 
  theme_minimal()

```

The residuals appear to be approximately normal except at the lower end of the range (and two points at the upper end of the range), which is showing signs of being short tailed. This is not ideal, but it is not a huge violation of normality.


Using `ggplot()`, make a plot of the jackknife residuals (obtained using the `rstudent()` function) on the y-axis and the fitted values of the model on the x-axis. Comment on whether this residual plot looks good. If it does not, indicate what the problem(s) is (are).

```{r}
#| label: residual_plot

ggplot(data, aes(x = fitted(mod_lm), y = rstudent(mod_lm))) + 
  geom_point() +
  geom_smooth(method = "loess", formula = y~x, se = FALSE) +
  labs(title = "Jackknife Residuals vs Fitted Values",
       x = "Fitted Values",
       y = "Jackknife Residuals") +
  theme_minimal()

```

The residual plot is showing signs of heteroskedasticity, as not only are the redisuals fanning out as the fitted values increase, but there is also a clear concentration of points between fitted values of 20 and 40. This is a problem because it indicates that the assumption of constant variance and the assumption of uniformly distributed errors are violated.

### Part e (6 points)
Since the residual plots do not look good, let's try to use weighted least squares. Following the code on page 118 and 119 of Notes 7, create a vector of weights and then fit a new model using weighted least squares. 

```{r}
#| label: wls_model

abs_resids <- abs(mod_lm$residuals)
fitted <- mod_lm$fitted.values
emod <- lm(abs_resids ~ fitted)
w <- 1/emod$fitted.values^2
mod_wls <- lm(compress_strength ~ ., data = data, weights = w)
summary(mod_wls)

```

### Part f (5 points)
Using `ggplot()`, make a plot of the jackknife residuals (obtained using the `rstudent()` function) on the y-axis and the fitted values of the weighted model on the x-axis. Comment on whether this residual plot looks good and whether this plot looks better than the residual plot in part d.

```{r}
#| label: wls_qq_plot

ggplot(data, aes(sample = rstudent(mod_wls))) + 
  geom_qq() +
  geom_qq_line() +
  labs(title = "QQ Plot of Jackknife Residuals (WLS)") + 
  theme_minimal()

```

```{r}
#| label: wls_residual_plot

ggplot(data, aes(x = fitted(mod_wls), y = rstudent(mod_wls))) + 
  geom_point() +
  geom_smooth(method = "loess", formula = y~x, se = FALSE) +
  labs(title = "Jackknife Residuals vs Fitted Values (WLS)",
       x = "Fitted Values",
       y = "Jackknife Residuals") +
  theme_minimal()

```

The QQ plot using the weighted least squares model is showing better signs of normality, but the normality assumption is still in question for the lower- and upper-most ends of the distribution, as it is now showing signs of being slightly skewed to the right. The residual plot is showing much better as the fanning has turned into a more consistant spread of residuals across the range of fitted values, more so up to fitted values of around 60, where the residuals tend to be negative more often. While there is still a slight concentration of points between fitted values of 20 and 40, it is not as bad as what we saw in the unweighted model. Overall, the residual plots for the weighted model are showing better signs of meeting the assumptions than the residual plots for the unweighted model.

### Part f (8 points)
Using the unweighted model from part c and the weighted model from part e, find and report both a confidence interval for the mean compressive strength and a prediction interval for the specific compressive strength for concrete that has a cement value of 300, a blast furnace slag of 90, a fly ash of 50, a water value of 200, a superplasticizer of 2.5, a coarse aggregate of 900, a fine aggregate of 600, and an age of 300 days. You can use the `predict()` function here and remember that you will need to find the specific weight for the given predictor variable values for the weighted intervals. You can follow along with the code used on page 121 of Notes 7.

Comment on how these intervals differ. Does the change make sense given the value of the fitted value?

```{r}
#| label: confidence_&_prediction_intervals

new_obs <- data.frame(cement = 300, blast_furnace_slag = 90, fly_ash = 50, water = 200, 
                           superplasticizer = 2.5, course_agg = 900, fine_agg = 600, age = 300)

predict(mod_lm, new_obs, interval = "confidence")

predict(mod_wls, new_obs, interval = "confidence")

predict(mod_lm, new_obs, interval = "prediction")

new_fit <- predict(mod_lm, new_obs)
new_weight <- 1/predict(emod, data.frame(fitted = new_fit))^2
predict(mod_wls, new_obs, interval = "prediction", weights = new_weight)

```

The OLS model predicts with 95% confidence that for all possible samples with the provided parameters, the true population average compresseion strength falls between 55.66444 - 63.87534 MPa. With the OLS model, if a new concrete sample is made with the provided parameters, it predicts with 95% confidence that the compressive strength of that particular sample will fall between 39.9145 - 97.00641 MPa. 

The WLS model predicts with 95% confidence that for all possible samples with the provided parameters, the true population average compresseion strength falls between 64.15589 - 72.82503 MPa. With the WLS model, if a new concrete sample is made with the provided parameters, it predicts with 95% confidence that the compressive strength of that particular sample will fall between 39.9145 - 97.06641 MPa.

We can see that both models have a wider prediction interval than their respective confidence intervals, which makes sense because the prediction interval is accounting for individual variability in the response variable, while the confidence interval is only accounting for variability in the mean variability in the response variable. 

We can see that the confidence interval for the WLS model is shifted to the right and is narrower than the confidence interval for the OLS model. Despite this shift, the prediction intervals for both models are pretty similar, with the WLS model having a slightly wider prediction interval. This makes sense because the fitted value for the WLS model is higher than the fitted value for the OLS model, which would shift the confidence interval to the right. The prediction intervals are similar because they are both accounting for individual variability in the response variable, which is not affected by the weighting in the WLS model.

### Part g (10 points)
Write a function called `predict_weighted` that takes three inputs: the **unweighted** model, a data frame containing information about the value(s) of the predictor variables we are using to predict, and the interval type (either "confidence" or "prediction"). This function should return the predicted value and the interval bounds for the specified interval type for weighted least squares. So, this function should compute the weights, obtain the weighted least squares model, find the specific weight for the new value, and then get the prediction and the interval. This function should work for any linear model (made with the `lm()` function) you give it, not just for this exact situation of this problem. So, you should not reference any data sets or variables specific to this concrete problem in the function.

Test this new function for the confidence and prediction intervals you made in part f. 

```{r}
#| label: predict_weighted_function

predict_weighted <- function(unweighted_model, new_data, interval_type) {
  abs_resids <- abs(mod_lm$residuals)
  fitted <- mod_lm$fitted.values
  emod <- lm(abs_resids ~ fitted)
  w <- 1/emod$fitted.values^2
  mod_wls <- lm(formula(unweighted_model), data = model.frame(unweighted_model), weights = w)

  new_fit <- predict(unweighted_model, new_data)
  new_weight <- 1/predict(emod, data.frame(fitted = new_fit))^2

  if (interval_type == "confidence") {
    return(predict(mod_wls, new_data, interval = "confidence"))
  } else if (interval_type == "prediction") {
    return(predict(mod_wls, new_data, interval = "prediction", weights = new_weight))
  } else {
    return("Invalid interval type. Please choose either 'confidence' or 'prediction'.")
  }
}

predict_weighted(mod_lm, new_obs, "confidence")
predict_weighted(mod_lm, new_obs, "prediction")
predict_weighted(mod_wls, new_obs, "none")

```

### Part h (5 points)
In this problem, based on the residual plots in part d, we found you the errors do not have constant variance. Another remedial measure for models with non-constant variance (heteroskedasticity) is to bootstrap the cases. 

Using the `boostrapping()` function in `math3190package`, run a bootstrap of the cases for at least 5,000 bootstrap samples on the model coefficients. Note that an unweighted model should be fit when bootstrapping.

```{r}
#| label: bootstrap_cases

bootcases <- function(resids, i) {
  lm(formula(mod_lm), data = model.frame(mod_lm)[i,])$coef
}
(boot_out <- bootstrapping(residuals(mod_lm), bootcases, 5000))
boot_ci(boot_out)

```

### Part i (5 points)
Compare the means and standard deviations for the bootstrapped coefficients along with the estimate and standard errors for the coefficients for both the weighted and unweighted models. Write a few sentences comparing them. If done correctly, the standard errors when bootstrapping should be larger. Why is that the case?

```{r}
#| label: comparison

(boot_out <- bootstrapping(residuals(mod_lm), bootcases, 5000))

summary(mod_lm)$coef

summary(mod_wls)$coef

```

Across all three models, the bootstrapped standard errors are consistently the largest. The OLS (unweighted) model generally yields the second-highest errors, with the exception of the age variable. This inflation in the bootstrapped errors is expected because the bootstrap captures the true variability and the "noise" caused by heteroskedasticity that the OLS and WLS formulas ignore. 
Regarding the coefficient estimates, the values remain relatively stable across all models for most variables. However, the superplasticizer and age variables show a noticeable shift in the WLS model compared to the OLS and bootstrapped results. This divergence occurs because WLS applies weights to minimize the influence of high-variance observations. If the relationship between these specific predictors and the response variable is different for the "more reliable" (low-variance) observations, the WLS estimates will naturally shift to reflect that subset of the data more heavily. 